Probability of causal contact between interstellar civilizations through MonteCarlo simulations

### 2020-05-05 submit_v03/manuscript.tex
starting reply to reviewer

\begin{abstract}
%
The abundance of intelligent civilizations in the Galaxy is a longstanding
question, often conceptualized as the problem of the lack of received
communication. Most efforts on the estimation of a number of civilizations are
focused on the Drake equation, although its factors are affected by large
uncertainties and it lacks a temporal nature.  Alternative approaches use
detailed numerical simulations of the galaxy and recipes for the stars and
planets formation rates and for the origin of life, and rely on highly
uncertain parameters. We present a statistical model for the abundance and
duration of civilizations, implemented through Monte Carlo simulations.  We
explore the hypothesis space of the model by a suite of simulations for the
emergence of communicating nodes, and analyze their
causal connections.  This is based on minimal assumptions and three
free parameters, with focus on the statistical properties of empirical
models instead of an interpretable model with variables to be determined by
observation. The analysis of the rate of causal contacts as a function of the
mean number of civilizations, the mean lifespan distribution and the maximum
distance a civilization can send signals is presented, and used as a framework
to discuss the spatial and temporal structure of a populated galaxy within
several scenarios. We find that, given the large distances involved,
causal contacts between civilizations are very rare. Additionally, the odds to
make a contact in a few years of monitoring are low for most models, with the
exception of those of a galaxy densely populated with long lived
civilizations. The probability of causal contacts increases with the
lifetime of civilizations much more significantly than with the number of
active civilizations for a given time window. Finally, we show that the
maximum probability of making a contact occurs when a civilization acquires
the required communication technology.
%
\end{abstract}

@@

The Drake equation \citep{drake_intelligent_1962} offers a very
helpful educated guess, a rational set of lenses --the factors in the
equation-- through which to look at future contacts with a
technologically advanced civilization in the Milky Way.

@@

receive an electromagnetic signal, using a number of factors that have
been widely discussed in the literature and whose estimates are
updated constantly.

@@

A comprehensive review and analysis of each term of the equation is

@@

Some optimistic estimates from the Drake equation contrast with the so
called Fermi paradox, a statement about the contradiction between the
apparent abundance of life in the Galaxy and its lack of evidence
\citep[e.g. ][]{hart_explanation_1975, brin_great_1983,

@@

There are many proposals aimed at solving this paradox, which make use
of statistical \citep{solomonides_probabilistic_2016,

@@

\citep{prantzos_joint_2013,
smith_broadcasting_2009} or reformulations

@@

We stress the fact that the only observation that can be stated with
certainty is that for the number of years projects for the Search of
Extraterrestrial Intelligences (SETI) have been working, we have not
received any signal using the technical possibilities and the
conditions established by SETI projects \citep{tarter_search_2001}.

@@

The lack of detection has also inspired to a number of alternative
proposals for new SETI strategies \citep{forgan_exoplanet_2017,
balbi_impact_2018, loeb_eavesdropping_2006, maccone_KLT_2010,
tarter_advancing_2009, enriquez_breakthrough_2017, loeb_relative_2016,
maccone_SETI_2011,  lingam_relative_2019, wright_theGsearch_2015,
maccone_SETI_2013, maccone_lognormals_2014, harp_application_2018,
forgan_possibility_2013, forgan_galactic_2017, funes_searching_2019}.

@@

The Drake equation is a key tool to organize the discussion around the
problem of the abundance of civilizations in the Galaxy
\citep{hinkel_interdisciplinary_2019}.

@@

However, the uncertainties in the factors of this equation make it
less prone to a formal application in order to define searching
strategies or to compute the actual number of extraterrestrial
intelligences.

@@

The optimistic estimations from the Drake equation also contrast with
the SETI initiatives, noticeable overshadowed by the fact that despite
decades of effort, not a single extraterrestrial intelligence, if any,
has been detected.

@@

\citet{prantzos_joint_2013} proposes a unified framework for a joint
analysis of the Drake equation and the Fermi paradox, concluding that
for sufficiently long-lived civilizations, colonization of the Galaxy
is the only reasonable option to gain knowledge about other life
forms.

@@

\citet{haqq-misra_drake_2017} discuss the dependence of the Drake
equation parameters on the spectral type of the host star and the time
since the Galaxy formed and examine trajectories for the emergence of
communicative civilizations over the history of the Galaxy.

@@

Some modifications to the original idea of the equation have tried to
imprint a stochastic nature, to propose a probabilistic approach, or
to consider the temporal structure which is missing in the equation.

@@

Temporal aspects of the distribution of communicating civilizations
and their contacts have also been explored by several authors
\citep{fogg_temporal_1987, forgan_spatiotemporal_2011,
balbi_impact_2018, balb_spatiotemporal_2018, horvat_impact_2011}, as
well as efforts on considering the stochastic nature of the Drake
equation \citep{glade_stochastic_2011}.

@@

The simulation approach has also been considered
\citep{forgan_evaluating_2015, vukotic_grandeur_2016,
murante_simulating_2015, forgan_numerical_2009, forgan_galactic_2017,
ramirez_new_2017}, although with a similar problem that consists on
the large number of parameters which are either unknown or largely
uncertain.

@@

\citet{grimaldi_signal_2017} consider a statistical model for the
probability of the Earth making contact with other intelligent
civilizations, taking into account the finite lifetime of signal
emitters, and based on the fractional volume occupied by all signals
reaching our planet.

In a recent work, \citep{balbi_impact_2018} use a statistical model to
analyze the occurrence of causal contacts between civilizations in the
Galaxy.
%
The author highlights the effect of evolutionary processes when
attempting to estimate the number of communicating civilizations that
might be in causal contact with an observer located on the Earth.
%
\citet{cirkovic_temporal_2004} also emphasize the lack of temporal
structure in the Drake equation.
%
\citet{bloetscher_using_2019} uses a probabilistic approach, though
still strongly motivated by the Drake equation, to obtain a
probabilistic measure for the number of civilizations in the Galaxy.
%
To that end, perform MonteCarlo Markov Chains over each factor of the
Drake equation, and combine the means to obtain a probabilistic
result.
%
It is worth mentioning that the author proposes a log-normal target
distribution to compute the posterior probabilities.
%
%This approach, which assigns probability distributions to the Drake
%factors.
%
This study concludes that there is a very low probability for Earth to
make contact with another extraterrestrial civilization.    
%
\citet{smith_broadcasting_2009} uses an analytical model to compute
the probabilities of contact between two randomly located
civilizations, and the waiting time for the first contact, assuming a
fixed maximum broadcasting distance.
% 
Using a different approach, \citet{balbi_impact_2018} investigate the
chance of communicating civilizations making causal contact within a
volume surrounding the location of the Earth, using a statistical
model to compute the occurrence of causal contacts.
%
The author stresses the fact that the causal contact requirement
involves the distance between civilizations, their lifespan and their
times of appearance.  This is important since the time it takes the
light to travel across the Galaxy is much lesser than its lifetime.
%
\citet{balbi_impact_2018} fix the total number of civilizations and
explore the parameter space that comprise three parameters, namely,
the distance to the Earth, the time of appearance and the lifespan of
the communicating civilization.
%
Each of these three variables are drawn from a random distribution.
The distribution of the distances results from a uniform distribution
of civilization within the plane of the Galaxy.
%
For the distribution of the characteristic time of appearance, the
author explores and exponential and a truncated Gaussian
distributions. For the distribution of the lifespans, the author
chooses an exponential distribution.
%
%The radius of influence is set to 1000 lyr.
%
\citet{balbi_impact_2018} finds that the fraction of communicating
civilizations vary with the choice of the statistical distribution for
the time of appearance. For all analyzed distributions, the fraction
corresponding to a mean lifetime of 10$^9$ years and for a maximum
radius for the detection of the signal of 1000 yr, is roughly 0.1.
%
In another probabilistic approach,
\citet{vukotic_astrobiological_2012} propose a probabilistic cellular
automata (PCA) modeling. In this framework, a complex system is
modeled by a lattice of cells which evolve at discrete time steps,
according to transition rules that take into account for each cell the
states of its neighbor cells.
%
The authors implement a PCA model to a network of cells which
represent life complexity on a 2-dimensional annular ring resembling
the Galactic Habitable Zone (GHZ). 
%
The authors set the GHZ between a minimum radius of 6 kpc and a
maximum radius of 10 kpc. Within this framework, the authors also make
several MonteCarlo simulations and analyze ensemble-averaged results.
%
This work aims at analyzing the evolution of life, although it does
not account for the network of causal contacts among technological
civilizations.

>>>>

\begin{abstract}
%
The abundance of intelligent civilizations in the Galaxy is a
   longstanding question, often conceptualized as the problem of the
   lack of received communication. Estimates of the number of
   civilizations are generally guided by the Drake equation, despite
   the large uncertainties in its factors and its lack of a temporal
   nature. Alternative approaches use detailed numerical simulations
   of the galaxy and recipes for the stars and planets formation
   rates and for the origin of life, and rely on uncertain parameters.
   We present a statistical model for the abundance and duration of
   civilizations implemented through Monte Carlo simulations. We
   explore the hypothesis space of the model by a suite of simulations
   to analyze the emergence of communicating nodes and their causal
   connections. We based this on minimal assumptions and three free
   parameters, with focus on the proposed statistical properties of
   empirical models. Our analysis of the dependence of the rate of
   causal contacts on the mean number of civilizations, the mean
   lifespan distribution and the maximum distance a civilization can
   send signals, is considered to discuss the spatial and temporal
   structure of a populated galaxy within several scenarios. We find
   that, given the enormous distances involved, causal contacts
   between civilizations are very rare. The odds to make a contact in
   a few years of monitoring are low for most models, except for those
   of a galaxy densely populated with long-lived civilizations. The
   probability of causal contacts increases with the lifetime of
   civilizations much more significantly than with the number of
   active civilizations for a time window. We show that the maximum
   probability of making a contact occurs when a civilization gains
   the required communication technology.
%
\end{abstract}

@@

The Drake equation  \citep{drake_intelligent_1962} provides a truly
helpful educated guess, a rational set of lenses –the factors in the
equation– through which to look at future contacts with
technologically advanced civilizations in the Milky Way. 

@@

receive an electromagnetic signal, using a collection of factors that
have been extensively discussed in the literature and whose estimates
are revised continually. 

@@

A comprehensive review and analysis of each term of the equation are

@@

Optimistic estimates from the Drake equation contrast with the
so-called Fermi paradox, which states the contradiction between the
expected abundance of life in the Galaxy and the lack of evidence for
it \citep[e.g. ][]{hart_explanation_1975, brin_great_1983,

@@

There are many propositions aimed at solving this paradox, which make
use of statistical \citep{solomonides_probabilistic_2016,

@@

\citep{prantzos_joint_2013, smith_broadcasting_2009} or reformulations

@@

We stress the fact that the only observation that can be stated with certainty is that for the number of years projects for the Search of Extraterrestrial Intelligences (SETI) have been working; we have received no signal using the technical possibilities and the conditions established by them \citep{tarter_search_2001}.

@@

The absence of detection has also motivated alternative ideas for new SETI strategies \citep{forgan_exoplanet_2017, balbi_impact_2018, loeb_eavesdropping_2006, maccone_KLT_2010, tarter_advancing_2009, enriquez_breakthrough_2017, loeb_relative_2016, maccone_SETI_2011, lingam_relative_2019, wright_theGsearch_2015, maccone_SETI_2013, maccone_lognormals_2014, harp_application_2018, forgan_possibility_2013, forgan_galactic_2017, funes_searching_2019}.

@@

The Drake equation is a key tool to organise the discussion about the problem of the (unknown) abundance of civilisations in the Galaxy \citep{hinkel_interdisciplinary_2019}.

@@

However, the uncertainties in the factors of this equation make it less prone to a formal study to define searching strategies or to compute the actual number of extraterrestrial intelligences.

@@

The optimistic estimations from the Drake equation also contrast with the SETI initiatives, noticeable overshadowed because despite decades of effort, not a single evidence for intelligent signals has been detected.

@@

\citet{prantzos_joint_2013} proposes a unified framework for a joint analysis of the Drake equation and the Fermi paradox, concluding that for sufficiently long-lived civilisations, colonisation of the Galaxy is the only reasonable option to gain awareness about other life forms.

@@

\citet{haqq-misra_drake_2017} discuss the dependence of the Drake equation parameters on the spectral type of the host stars and the time since the Galaxy formed and examine trajectories for the emergence of communicative civilisations.

@@

Some modifications to the original idea of the equation have tried to imprint a stochastic nature, to propose a probabilistic approach, or to consider the temporal structure missing in the equation.

@@

Temporal aspects of the distribution of communicating civilisations and their contacts have also been explored by several authors \citep{fogg_temporal_1987, forgan_spatiotemporal_2011, balbi_impact_2018, balb_spatiotemporal_2018, horvat_impact_2011}, as well as efforts on considering the stochastic nature of the Drake equation \citep{glade_stochastic_2011}.

@@

The literature about this topic brings a number of works that consider numerical simulations
\citep{forgan_evaluating_2015, vukotic_grandeur_2016, murante_simulating_2015, forgan_numerical_2009, forgan_galactic_2017, ramirez_new_2017}.

@@

Although this approach does not rely on observed quantities that could be modified with time, it also depend on the definition of a number of parameters which are unknown or uncertain.
\citet{grimaldi_signal_2017} consider a statistical model for the probability of the Earth contacting other intelligent civilisations, taking into account the finite lifetime of signal emitters, and based on the fractional volume occupied by all signals reaching our planet.

In a recent work, \citep{balbi_impact_2018} use a statistical model to analyse the occurrence of causal contacts between civilisations in the Galaxy.
%
The author highlights the effect of evolutionary processes when attempting to estimate the number of communicating civilisations that might be in causal contact with an observer on the Earth.
%
\citet{cirkovic_temporal_2004} also emphasise the lack of temporal structure in the Drake equation.
%
\citet{bloetscher_using_2019} uses a probabilistic approach, though still motivated by the Drake equation, to obtain a probabilistic measure for the number of civilisations in the Galaxy.
%
To that end, the authors do Monte-Carlo Markov Chains over each factor of the Drake equation, and combine the means to reach a probabilistic result.
%
It is worth mentioning that the author proposes a log-normal target distribution to compute the posterior probabilities.
%
This study concludes that there is a tiny probability for Earth to contact another extraterrestrial civilization.    
%
\citet{smith_broadcasting_2009} uses an analytical model to gauge the probabilities of contact between two randomly located civilisations, and the waiting time for the first contact, assuming a fixed maximum broadcasting distance.
% 
Adopting a novel approach, \citet{balbi_impact_2018} investigate the chance of communicating civilisations making causal contact within a volume surrounding the location of the Earth, using a statistical model to compute the occurrence of causal contacts.
%
The author stresses the fact that the causal contact requirement involves the distance between civilisations, their lifespan and their times of appearance. This is important since the time it takes the light to travel across the Galaxy is much lesser than its lifetime.
%
\citet{balbi_impact_2018} fix the total number of civilisations and explore the parameter space that comprise three parameters, namely, the distance to the Earth, the time of appearance and the lifespan of the communicating civilization.
%
Each of these three variables are drawn from a random distribution. The distribution of the distances results from a uniform distribution of civilizations within the plane of the Galaxy.
%
For the distribution of the characteristic time of appearance, the author explores an exponential and a truncated Gaussian distributions. For the distribution of the lifespans, the author chooses an exponential distribution.

%-------------------------------------------------------------------
%%%%%%%%%%%%%%%%%%%%%%%%%% FIRST CHECK %%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%-------------------------------------------------------------------

%%%%% SIGUE PUNTO Y SEGUIDO
%
%The radius of influence is set to 1000 lyr.
%
\citet{balbi_impact_2018} finds that the fraction of communicating
civilizations vary with the choice of the statistical distribution for
the time of appearance. For all analyzed distributions, the fraction
corresponding to a mean lifetime of 10$^9$ years and for a maximum
radius for the detection of the signal of 1000 yr, is roughly 0.1.
%
In another probabilistic approach,
\citet{vukotic_astrobiological_2012} propose a probabilistic cellular
automata (PCA) modeling. In this framework, a complex system is
modeled by a lattice of cells which evolve at discrete time steps,
according to transition rules that take into account for each cell the
states of its neighbor cells.
%
The authors implement a PCA model to a network of cells which
represent life complexity on a 2-dimensional annular ring resembling
the Galactic Habitable Zone (GHZ). 
%
The authors set the GHZ between a minimum radius of 6 kpc and a
maximum radius of 10 kpc. Within this framework, the authors also make
several MonteCarlo simulations and analyze ensemble-averaged results.
%
This work aims at analyzing the evolution of life, although it does
not account for the network of causal contacts among technological
civilizations.
